---
title: "ToOoOlTiPs: An R Package for Customizable Tooltips in Interactive Graphics"
date: "26-06-2023"
abstract: >
  An abstract of less than 150 words.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Quietest Quokka
    affiliation: Calvin University
    address:
    - Department of Letter Q
    - Somewhere, Australia
    url: https://www.britannica.com/animal/quokka
    orcid: 0000-1721-1511-1101
    email:  qquo@ulm.edu
  - name: Bounciest Bilby
    affiliation:
    - University of Little Mates
    - University of Aussie Animals
    address:
    - Department of Letter Q, Somewhere, Australia
    - Department of Marsupials, Somewhere, Australia
    url: https://www.britannica.com/animal/bilby
    email: bbil@ulm.edu
    orcid: 0000-0002-0912-0225
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tagtools)
library(plotly)
library(stats)
library(ggplot2)
```

# Introduction

Interactive data graphics provides plots that allow users to interact them. One of the most basic types of interaction is through tooltips, where users are provided additional information about elements in the plot by moving the cursor over the plot.

This paper will first review some R packages on interactive graphics and their tooltip implementations. A new package \CRANpkg{ToOoOlTiPs} that provides customized tooltips for plot, is introduced. Some example plots will then be given to showcase how these tooltips help users to better read the graphics. 

# Background

Some packages on interactive graphics include \CRANpkg{plotly} [@plotly] that interfaces with Javascript for web-based interactive graphics, \CRANpkg{crosstalk} [@crosstalk] that specializes cross-linking elements across individual graphics.  The recent R Journal paper \CRANpkg{tsibbletalk} [@RJ-2021-050] provides a good example of including interactive graphics into an article for the journal. It has both a set of linked plots, and also an animated gif example, illustrating linking between time series plots and feature summaries.

# Customizing tooltip design with \pkg{ToOoOlTiPs}

\pkg{ToOoOlTiPs} is a packages for customizing tooltips in interactive graphics, it features these possibilities.

# Processing

For calculations pertaining to data collected, we have a host of functions available to process the data for different purposes. As a standout of these functions, we have find_dives. This function is used to find the time cues for the start and end of either dives in a depth record or flights in an altitude record. The simulated quantities used to do this are: 

p: A depth or altitude time series (a sensor data list or  a vector) in meters.
sampling_rate: The sampling rate of the sensor data in Hz (samples per second).
mindepth: The threshold in meters at which to recognize a dive or flight. Dives shallow or flights lower than mindepth will be ignored.
surface (optional): The threshold in meters at which the animal is presumed to have reached the surface. Default value is 1. A smaller value can be used if the dive/altitude data are very accurate and you need to detect shallow dives/flights.
findall (optional) When TRUE, forces the algorithm to include incomplete dives at the start and end of the record. Default is FALSE which only recognizes complete dives.

Find_dives creates a data frame with one row for each dive or flight found. The columns of T are: 
start (time in seconds of the start of each dive/flight)
end (time in seconds of the start of each dive/flight)
max (maximum depth/altitude reached in each dive/flight) 
tmax(time in seconds at which the animal reaches the max depth/altitude).

Using the dataset beaked_whale from the package as BW and
T as find_dives(p = BW$P$data, sampling_rate = BW$P$sampling_rate, mindepth = 5, surface = 2, findall = FALSE), the result produced are:


```{r}
BW <- beaked_whale
find_dives(p = BW$P$data, sampling_rate = BW$P$sampling_rate, mindepth = 5, surface = 2, findall = FALSE)
```

Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:penguins-tab-interactive)', '\\@ref(tab:penguins-tab-static)'))` prints at the first few rows of the `penguins` data:

```{r penguins-tab-interactive, eval = knitr::is_html_output(), layout = "l-body-outset"}
knitr::kable(head(penguins), format = "html", caption = "A basic table")
```

```{r penguins-tab-static, eval = knitr::is_latex_output()}
knitr::kable(head(penguins), format = "latex", caption = "A basic table") %>% 
  kableExtra::kable_styling(font_size = 7)
```   

Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(fig:penguins-plotly)', '\\@ref(fig:penguins-ggplot)'))` shows an `r ifelse(knitr::is_html_output(), 'interactive', '')` plot of the penguins data, made using the `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\CRANpkg{plotly}', '\\CRANpkg{ggplot2}'))` package. 

```{r penguins-plotly, echo = TRUE, fig.height = 5, fig.cap="A basic interactive plot made with the plotly package on palmer penguin data. Three species of penguins are plotted with bill depth on the x-axis and bill length on the y-axis. When hovering on a point, a tooltip will show the exact value of the bill depth and length for that point, along with the species name.", include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.alt = "A scatterplot of bill length against bill depth, both measured in millimetre. The three species are shown in different colours and loosely forms three clusters. Adelie has small bill length and large bill depth, Gentoo has small bill depth but large bill length, and Chinstrap has relatively large bill depth and bill length."}
p <- penguins %>% 
  ggplot(aes(x = bill_depth_mm, y = bill_length_mm, 
             color = species)) + 
  geom_point()
ggplotly(p)
```

```{r penguins-ggplot, echo = TRUE, fig.height = 5, fig.cap="A basic non-interactive plot made with the ggplot2 package on palmer penguin data. Three species of penguins are plotted with bill depth on the x-axis and bill length on the y-axis. Visit the online article to access the interactive version made with the plotly package.", include=knitr::is_latex_output(), eval=knitr::is_latex_output()}
penguins %>% 
  ggplot(aes(x = bill_depth_mm, y = bill_length_mm, 
             color = species)) + 
  geom_point()
```

# Summary

We have displayed various tooltips that are available in the package \pkg{ToOoOlTiPs}. 
