---
title: "Read in data from SMRT tag deployment and save calibrated data in netCDF format"
format: typst
editor: source
---

```{r}
#| label: setup
#| include: false
library(tagtools)
# AARGH it has a wav object classlibrary(tuneR) # for reading wav files - make it a tagtools dependency or suggestion if it works well.
library(wav)
```

```{r}
#| label: paths-metadata

# note: functions in the smrt_r folder could/will be added to the R tagtools package when ready.
data_dir <- "/Volumes/Boomin/Zica-20191117-195993"
depid <- "Zica-20191117-195993"
sm_dir <- paste0(datapath, "/SMFiles")
gps_file <- paste0(datapath, "/WCfiles/DAP580/Zica-20191117-195993-19U1753-FastGPS.csv")
archive_file <- paste0(datapath, "/WCfiles/DAP580/Zica-20191117-195993-19U1753-Archive")
tz <- "UTC"
species <- "zc"

info <- make_info(depid = depid, tagtype = "SMRT", species = species, owner = "gss")
info$project_name <- "Freakin Beakin"
info$dephist_device_tzone = "UTC"

mag_strength = 45.2689; 
# expected field strength in uT 
# obtained from https://www.ngdc.noaa.gov/geomag/calculators/magcalc.shtml#igrfwmm
# you can also get heading declination/inclination from this site too
```

```{r}
# read_smrt() will call the three main functions to read the different pieces of SMRT data:
# read_smrt_gps() for GPS data
# read_smrt_archive() for data from WC board
# read_smrt_sm() for data from SM board
```


# Notes: Expected DTAG problems if using these R SMRT tools with DTAGs

- d3builtincal.m is not implemented in R, so functions to read swv files don't call it and don't make this correction. get_sm_config() also does not look for a CAL structure within the xml files. These would need to be added for d3 compatibility, I think. The "builtincal" m-file says, "This is called by d3parseswv and d3getswv to apply calibrations to the temperature and pressure channels for tags with I2C pressure sensors."
- cuetabs and file name lists ARE NOT stored in a temp directory. The would need to be regenerated each time they are used...or alternatively stored in SA (sound archive) sensor data list or in the deployment level info list, within an nc file!
