---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tagtools

<!-- badges: start -->
``` {r, echo = FALSE}
usethis::use_cran_badge()
usethis::use_lifecycle_badge("stable")
```
[![R-CMD-check](https://github.com/animaltags/tagtools_r/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/animaltags/tagtools_r/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/tagtools)](https://CRAN.R-project.org/package=tagtools)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

This repository contains an R package with the R version of the animaltag tool kit. High-resolution movement-sensor tags typically include accelerometers to measure body posture and sudden movements or changes in speed, magnetometers to measure direction of travel, and pressure sensors to measure dive depth in aquatic or marine animals. The sensors in these tags usually sample many times per second. Some tags include sensors for speed, turning rate (gyroscopes), and sound. This package provides software tools to facilitate calibration, processing, and analysis of such data. Tools are provided for: data import/export; calibration (from raw data to calibrated data in scientific units); visualization (for example, multi-panel time-series plots); data processing (such as event detection, calculation of derived metrics like jerk and dynamic acceleration, dive detection, and dive parameter calculation); and statistical analysis (for example, track reconstruction, a rotation test, and Mahalanobis distance analysis).

## Installation

You can install the current CRAN version of tagtools with:

``` r
install.packages("tagtools")
```

You can install the development version of tagtools from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("animaltags/tagtools_r")
```

## Further background

Documentation and vignettes are available at: <https://animaltags.github.io/tagtools_r/>

This material is based upon research supported by the United States Office of Naval Research under Award Number N00014-16-1-3089.
