---
title: "SMRT netCDF file and filtering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SMRT netCDF file and filtering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tagtools)
```

# Preparation of Data & Metadata

## Obtain data
This vignette presents analysis of a dataset that is publicly available, but is much too large to include in an R package. 

If you would like to reproduce this example yourself, using the example dataset, first download the data from **NEED LINK HERE** and make a note of where on your computer it is stored.

## File formats

SMRT tag data includes multiple file types:

- .xml files containing metadata about the tag settings
- .csv files containing GPS data and sensor data (often wet/dry, temperature, pressure and others)
- .swv files, which are files that follow the .wav file format standards and contain sensor data (often from accelerometer, magnetometer, and others)
- .wav files with acoustic data
- .wavt files (sometimes present) with data about the timing of data recording and any gaps in recording

The `tagtools` package includes functions to read in data from all these file types and combine it for storage in an archival netCDF (.nc) file.

The .nc file includes data from the non-acoustic sensors, metadata about the deployment, and metadata about the acoustic recordings. (The acoustic data are not incorporated into the .nc file directly since they are huge, and since they will often be processed and analyzed separately and using tools that accept and expect data in .wav format.)

## Directory structure

While it should be possible to read in your SMRT data files using `tagtools` functions as long as you have all the data files and know where they are stored, a typical directory structure for SMRT data would include one main directory per deployment, and within that a sub-folder (perhaps called `SMfiles`) for the files generated by the "SM" (sound and motion) board -- these are the files that are "unpacked" from raw .dtg files off the tag into .wav, .swv, .wavt, .xml, and .csv files. 

## Metadata

Before we begin, in addition to knowing where our data files are stored, we will need a bit of metadata.

First, specify the deployment ID (`depid` for short). There is no requirement for this to follow a set format, but it should match the data file names. For example, data from tag "Zica-20200106-195994" would include files like "Zica-20200106-195994001.swv" and "Zica-20200106-195994001.xml".

```{r}
depid <- "Zica-20200106-195994"
```

We also need to specify where the data is stored -- minimally a main data directory, and perhaps (especially if the folder with the SM files is not called "SMfiles") the directory where SM files are located.

```{r}
my_data_dir <- "tagdata/Zica-20200106-195994"
my_sm_dir <-file.path(my_data_dir, "SMfiles")
```

Finally, it might be wise to pull together the deployment-level metadata now. 

Some information (like the device ID, the exact recording start time, and sensors present, for example) can and probably should be obtained directly from the data files. Other information is not stored in the tag's data files, such as the name and contact information of the data owner, and the species on which the tag was deployed; you will have to add this crucial information.

An initial list of metadata from the tag data files, in a format suitable for storage in the .nc file, can be created by a call to `make_info()`:

```{r}
my_info <- make_info(depid = depid, 
                     tagtype = "SMRT", 
                     species = "zc" # for Ziphius cavirostris (case sensitive)
                     ) 
```

The `tagtools` package includes .csv files with basic information about some biologging tags, potential tag data owners (file name "researchers.csv"), and species (file name "species.csv"). These are of course nowhere near comprehensive, and you will likely need to edit them to add your own information. You can find out where the files are stored (so you can edit them if needed) via a call to `system.file()` with the filename of interest:

```{r}
system.file("extdata", 
            "species.csv", 
            package = "tagtools")
```

Alternatively, you can create an emptier metadata structure and fill it in by hand. The .csv files are intended as a shortcut to make it easy to fill in *your* lab and project's common entries consistently.

```{r}
my_info <- make_info(depid = depid,
                     tagtype = "SMRT")
```
`my_info` is a list of metadata, containing plenty of empty fields you might choose to fill in now.

```{r}
str(my_info)
```

```{r}
my_info$animal_species_common <- "Goose-beaked whale"
my_info$animal_species_science <- "Ziphius cavirostris"
my_info$project_name <- "Super Science Project"
my_info$provider_name <- "Greg Schorr; Erin Falcone"
my_info$provider_details <- "Marine Ecology and Telemetry Research" 
my_info$provider_email <- "gschorr@marecotel.org; efalcone@marecotel.org"
```

Key fields to fill in manually are the ones that identify the tagged species, and give names and contact details for the data owners. Others will be filled in automatically as the data files are read, such as the device ID and recording start time.

The info fields that are necessary -- and their names -- may evolve as standards for archiving high-resolution multi-sensor tags evolve, and once such standards emerge `tagtools` will do our best to update accordingly.

# Read SMRT data

The function `read_smrt()` reads in data from a SMRT tag and saves an initial netCDF file ("initial" because more calibration and processing may be needed later).

The function has *many* inputs to allow for finer control and advanced or special-case use, but the key inputs are just the ones we have already defined. If you want to save the output .nc file somewhere *other* than the `data_dir`, you may also want to define the input `nc_dir` to specify where it should be saved.

Depending on dataset size, reading/processing/saving a SMRT dataset may take time; for deployments lasting days to about a week with some sensors sampling at 100Hz or more, single- to double-digit minutes of waiting time would not be impossible. 

You can set input `quiet = TRUE` to suppress (most of the) copious output about progress, although you may still get some messages and warnings related to any data gaps found and filled by the `tagtools`.

```{r}
my_nc_file_name <- read_smrt(data_dir = my_data_dir,
                             info = my_info,
                             depid = depid,
                             sm_dir = my_sm_dir,
                             quiet = FALSE)
```






