---
title: "SMRT netCDF file and filtering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SMRT netCDF file and filtering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tagtools)
```

# Preparation of Data & Metadata

## Obtain data

This vignette presents analysis of a dataset that is publicly available, but is much too large to include in an R package.

If you would like to reproduce this example yourself, using the example dataset, first download the data from **NEED LINK HERE** and make a note of where on your computer it is stored.

## File formats

SMRT tag data includes multiple file types:

- .xml files containing metadata about the tag settings
- .csv files containing GPS data and sensor data (often wet/dry, temperature, pressure and others)
- .swv files, which are files that follow the .wav file format standards and contain sensor data (often from accelerometer, magnetometer, and others)
- .wav files with acoustic data
- .wavt files (sometimes present) with data about the timing of data recording and any gaps in recording

The `tagtools` package includes functions to read in data from all these file types and combine it for storage in an archival netCDF (.nc) file.

The .nc file includes data from the non-acoustic sensors, metadata about the deployment, and metadata about the acoustic recordings. (The acoustic data are not incorporated into the .nc file directly since they are huge, and since they will often be processed and analyzed separately and using tools that accept and expect data in .wav format.)

## Directory structure

While it should be possible to read in your SMRT data files using `tagtools` functions as long as you have all the data files and know where they are stored, a typical directory structure for SMRT data would include one main directory per deployment, and within that a sub-folder (perhaps called `SMfiles`) for the files generated by the "SM" (sound and motion) board -- these are the files that are "unpacked" from raw .dtg files off the tag into .wav, .swv, .wavt, .xml, and .csv files.

## Metadata

Before we begin, in addition to knowing where our data files are stored, we will need a bit of metadata.

First, specify the deployment ID (`depid` for short). There is no requirement for this to follow a set format, but it should match the data file names. For example, data from tag "Zica-20200106-195994" would include files like "Zica-20200106-195994001.swv" and "Zica-20200106-195994001.xml".

```{r}
depid <- "Zica-20200106-195994"
```

We also need to specify where the data is stored -- minimally a main data directory, and perhaps (especially if the folder with the SM files is not called "SMfiles") the directory where SM files are located.

```{r}
my_data_dir <- "tagdata/Zica-20200106-195994"
my_sm_dir <-file.path(my_data_dir, "SMfiles")
```

Finally, it might be wise to pull together the deployment-level metadata now.

Some information (like the device ID, the exact recording start time, and sensors present, for example) can and probably should be obtained directly from the data files. Other information is not stored in the tag's data files, such as the name and contact information of the data owner, and the species on which the tag was deployed; you will have to add this crucial information.

An initial list of metadata from the tag data files, in a format suitable for storage in the .nc file, can be created by a call to `make_info()`:

```{r}
my_info <- make_info(depid = depid, 
                     tagtype = "SMRT", 
                     species = "zc" # for Ziphius cavirostris (case sensitive)
                     ) 
```

The `tagtools` package includes .csv files with basic information about some biologging tags, potential tag data owners (file name "researchers.csv"), and species (file name "species.csv"). These are of course nowhere near comprehensive, and you will likely need to edit them to add your own information. You can find out where the files are stored (so you can edit them if needed) via a call to `system.file()` with the filename of interest:

```{r}
system.file("extdata", 
            "species.csv", 
            package = "tagtools")
```

Alternatively, you can create an emptier metadata structure and fill it in by hand. The .csv files are intended as a shortcut to make it easy to fill in *your* lab and project's common entries consistently.

```{r}
my_info <- make_info(depid = depid,
                     tagtype = "SMRT")
```

`my_info` is a list of metadata, containing plenty of empty fields you might choose to fill in now.

```{r}
str(my_info)
```

```{r}
my_info$animal_species_common <- "Goose-beaked whale"
my_info$animal_species_science <- "Ziphius cavirostris"
my_info$project_name <- "Super Science Project"
my_info$provider_name <- "Greg Schorr; Erin Falcone"
my_info$provider_details <- "Marine Ecology and Telemetry Research" 
my_info$provider_email <- "gschorr@marecotel.org; efalcone@marecotel.org"
# my_info$dephist_deploy_location_lat
# my_info$dephist_deploy_location_lon # add deployment location here too
```

Key fields to fill in manually are the ones that identify the tagged species, and give names and contact details for the data owners. Others will be filled in automatically as the data files are read, such as the device ID and recording start time.

The info fields that are necessary -- and their names -- may evolve as standards for archiving high-resolution multi-sensor tags evolve, and once such standards emerge `tagtools` will do our best to update accordingly.

# Read SMRT data

The function `read_smrt()` reads in data from a SMRT tag and saves an initial netCDF file ("initial" because more calibration and processing may be needed later).

The function has *many* inputs to allow for finer control and advanced or special-case use, but the key inputs are just the ones we have already defined. If you want to save the output .nc file somewhere *other* than the `data_dir`, you may also want to define the input `nc_dir` to specify where it should be saved.

Depending on dataset size, reading/processing/saving a SMRT dataset may take time; for deployments lasting days to about a week with some sensors sampling at 100Hz or more, single- to double-digit minutes of waiting time would not be impossible.

You can set input `quiet = TRUE` to suppress (most of the) copious output about progress, although you may still get some messages and warnings related to any data gaps found and filled by the `tagtools`.

```{r}
my_nc_file_name <- read_smrt(data_dir = my_data_dir,
                             info = my_info,
                             depid = depid,
                             sm_dir = my_sm_dir,
                             quiet = FALSE)
```

If we want to continue working with the data, we can read it into R with the `tagtools` function `load_nc()`:

```{r}
smrt_data <- load_nc(my_nc_file_name)
```

# Calibration

The SMRT tag data (and metadata) has now been out of its tag-specific raw data files and stored in netCDF format.

But...there's more processing still to do! It's crucial to calibrate the data to ensure it's in correct engineering units for further analysis.

Before we begin, we collect information about sensor settings and calibration constants.

```{r}
CAL <- sm_get_cal(sm_dir = my_sm_dir)
```

(You can peek inside the `CAL` list to see what info it contains, gathered from the deployment .xml files, device bench calibration (if any), and manufacturer sensor specifications.)

```{r}
str(CAL)
```

## Cropping

It is common to archive (and analyze) only the data collected while the tag was on the animal. (Often tags continue recording after they release from the animal.)

One way to determine the on-animal time is by examination of the depth profile and acceleration data. We can use `plott()` to make a quick graph. Working with your own data, you may want to use `plott(..., interactive = TRUE)` to get a plot where you can easily zoom and mouse-over points of interest to obtain time estimates.

```{r}
plott(X = list(depth = smrt_data$depth, A = smrt_data$acceleration))
```

In our case, during the second, shallower dive just before 1.4 hours, there's a big spike in the acceleration data and a concurrent jog in the depth trace (maybe resulting from a sudden movement or impact). Then, the acceleration data is very static until the tag reaches the surface, where it gets noisier (but with each axis's data maintaining a steady average value, unlike when the tag was on the swimming whale). This pattern is characteristic of a tag coming off, then floating to and at the surface.

We want to keep the data starting at the 0th second of the recording, and ending at about 4886 seconds since recording start (1.36 hours).

So we use `crop_all()` with `tcues` (start- and end-**t**ime **cues**) `c(0, 4886)`.

```{r}
smrt_data <- crop_all(tcues = c(0, 4886), smrt_data)
```

We can also note that each sensor data list within `smrt_data` now has entries "crop" and "crop_units,"

```{r}
smrt_data$depth$crop
smrt_data$depth$crop_units
```

and their "history" entries also note that they have been cropped with `crop_to()` (which `crop_all()` applies to each sensor).

```{r}
smrt_data$depth$history
```

TODO: need to write/implement...

- ~~fix_offset_pressure~~
- ~~d4readattr~~
- ~~auto_cal_acc~~ 
  - NOTE: when and where does the CAL get saved? For DTAG? for SMRT? We may need to save it...current tools I think do not, unless d4findcal looks for deployment specific files (and if it does: when do those get saved?)
- auto_cal_mag 

## Depth (Pressure sensor) Data

For SMRT tags, the "raw" pressure data has already undergone a manufacturer calibration to convert to engineering units (m depth), including a correction for the temperature-sensitivity of the pressure sensor. So, for SMRT data, the `fix_pressure()` function doesn't generally improve data calibration. However, it may still be beneficial to apply `fix_offset_pressure()` to adjust surfacings to average closer to the nominal sea-surface depth value of 0.

Need to write fix_offset_pressure()!

```{r}

```

## Accelerometer Data

Will have to translate function d4readattr to get CAL!

Then write auto_cal_acc to apply the cal

```{r}

```

## Magnetometer Data

Requires CAL and M and internal temp.

```{r}
#| eval: false

# matlab version for reference
imag = ~cellfun('isempty', regexp(chan_descrips, 'magnet'));
imagt = ~cellfun('isempty', regexp(chan_descrips, 'temper'));
imag = imag & ~imagt ;
M = sens_struct([X.x{find(imag)}], X.fs(find(imag, 1)), depid, 'mag') ;
Td = interp2length(T1, M);
Pd = interp2length(P1, M);
cal = CAL.MAG ;
cal.poly = [1 0;1 0;1 0] ;
fprintf('Calibrating magnetometer data...\n')
if length(X.x)==7 % if there is an internal temperature channel -- could also search for "mag temp" in channel name. Deviation should improve again
    Tint = sens_struct(X.x{7}*4096+15, X.fs(7), depid, 'temp');
    Tint.name = 'Tint' ;
    Tint.description = 'internal temperature from LIS44' ;
    [M,cal1] = auto_cal_mag(M, cal, mag_strength, [Td.data Td.data Tint.data Pd.data/100], [60 600 0 0]);
    M.cal_tsrc = 'T,T,Tint,P/100...Tint.data(1:25) = Tint.data(26)' ;
else  % otherwise, approximate internal temperature by slowing down external temperature
    % make Pd.data with no neg to avoid complex numbers
    Tint = [];
    Pd_pos = Pd.data;
    Pd_pos(Pd.data < 0) = 0;
    [M,cal1] = auto_cal_mag(M, cal, mag_strength, [Td.data Td.data Td.data Pd.data/100 sqrt(Pd_pos/100)], [300 60 800 0 0]);
    M.cal_tsrc = 'T,T,T,Tsq,P/100,(T-20).^2' ;
end
```

## Frame-of-Reference Conversion

The tag (of course) records data in its own 3D frame of reference, with the 3D axis orientation just depending on how tri-axial sensors like accelerometers or magnetometers are affixed in the tag.

We (of course) care more about data in an animal-specific frame of reference: we would rather that canonical "forward" mean "in front of the animal." But tags deployed on free-swimming animals via poles, drones or launchers are not always placed so that the tag's axes ("tag frame") line up perfect with the animal's ("animal frame"). We would like to use the data (and some assumptions about how the animal moves) to determine how best to rotate the data to convert it from tag frame to animal frame.

There is potential to use machine-learning methods to automate this process.

Currently, the `tagtools` package also contains two functions that can be used (with oversight from the user!) to help figure it out, for tags deployed on cetaceans (like whales and dolphins).

- `prh_predictor1()` is suited to animals that "log" at the surface: that is, the animal rests horizontally at the surface (at least on average) and dives steeply away from the surface without an initial roll. Sperm whales might be one example.
- `prh_predictor2()` is suited to animals that make short dives between respirations. Beaked whales might be one example.

*Either actually show the whole process here, or else punt it to existing tutorial materials and just show the conversion.*

# (Re) Save netCDF File

Now that we have completed some calibration and data checks, we're ready to save an .nc file with the cropped, calibrated dataset.

```{r}
save_nc(smrt_data, file = file.path(my_data_dir, paste0(depid, "-cal.nc"))
```

We can use `load_nc()` to read in the data from this file for further analysis.

# Next Vignettes

When this one is done, we need:

- Filtering (and [horrors of] Aliasing)
  - show body_axes data with and without decimation-induced aliasing
  - recap the case of the dancing penguins
  - Report on timing of filtering for large datasets with and without go_slow. In text also report Matlab stats.
- Add CATS example? Or not...if so need to ensure that the extra roll \*-1 is fixed in read_cats() functions and need to decide what to do about magnetometer sampling rates. Also need to coordinate re: example dataset that can be shared.
